Characterizing the Transcriptome of Turkey Hemorrhagic Enteritis Virus in a Turkey B-cell Line

### 2024-02-25 manuscript_thev_transcriptome.Rmd
modified acknowledgment section
implemented prof edits to manuscript to submit

---
title: Characterizing the Transcriptome of Turkey Hemorrhagic Enteritis Virus
output:
  officedown::rdocx_document:
    reference_docx: style_word_output.docx
  pdf_document: default
geometry: margin=1in
font-size: 11pt
header-includes:
  - \usepackage{helvet}
  - \renewcommand*\familydefault{\sfdefault}
  - \usepackage{setspace}
  - \doublespacing
  - \usepackage[left]{lineno}
  - \linenumbers
  - \usepackage{booktabs}
  - \usepackage{longtable}
  - \usepackage{array}
  - \usepackage{multirow}
  - \usepackage{wrapfig}
  - \usepackage{float}
  - \usepackage{colortbl}
  - \usepackage{pdflscape}
  - \usepackage{tabu}
  - \usepackage{threeparttable}
  - \usepackage{threeparttablex}
  - \usepackage[normalem]{ulem}
  - \usepackage{makecell}
  - \usepackage{xcolor}
bibliography: transcriptome_refs.bib
csl: asm.csl
editor_options:
  chunck_output_type: console
---

@@

Abraham Quaye^1\*^, Brett Pickett^\*^, Joel S. Griffitts^\*^, Bradford K. Berges^\*^, Brian D. Poole${^\dagger}$^\*^

@@

```{r call_libs, echo=FALSE,include=FALSE, message=FALSE, warning=FALSE}
library(knitr)
library(tidyverse)
library(tibble)
library(scales)
library(flextable)
library(english)
library(rticles)
opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
knitr::opts_chunk$set(tab.cap.pre = "", tab.cap.sep = "")

inline_hook <- function(x) {
  if(is.numeric(x)){
    formatted <- format(x, big.mark = ",")
  }else{formatted <- x}
  paste0("**", formatted, "**")
}
knit_hooks$set(inline = inline_hook)
```

@@

We have performed an RNA-sequencing experiment characterizing the transcriptome of turkey hemorrhagic enteritis virus (THEV) for the first time, yielding the only insight into THEV's gene expression patterns. Previously, THEV's genome had been predicted to encode 23 open reading frames (ORFs). In this work we identified 29 transcripts from our RNA-seq data all of which consisted of novel exons albeit some exons matched the predicted ORFs. The three predicted splice junctions were also corroborated by our data. We performed PCR amplification of THEV cDNA and cloned the PCR products, and Sanger sequencing was used to validate all identified splice junctions. During validation, we identified 5 additional transcripts some of which were further validated by 3'RACE data. Thus, the transcriptome of THEV consists of 34 unique transcripts with the coding capacity for all predicted ORFs. However, we found 6 predicted ORFs (ORF1, E3, 33K, ORF8, IVa2, and protease) to be truncated predictions as either an in-frame upstream start codon was identified or additional coding exons were found. We also identified 3 predicted ORFs with longer or shorter isoforms, and 7 novel unpredicted ORFs that could be encoded by some transcripts; albeit it is beyond the scope of this manuscript to investigate whether they get translated. Similar to other adenoviruses, THEV also produce multiple distinctly spliced transcripts that code for the same protein across its genome. Also, our data shows that all THEV transcripts are spliced, and organized in five transcription units under the control of their cognate promoter. However, our data suggests that THEV's temporal regulation may be different from other adenoviruses. Over 2,300 unique splice junctions were found across the genome, mostly at low levels. This low-level use of broad alternative splicing patterns is thought to enable the virus to maximize its coding potential in an evolving environment. 
\newpage

@@

Adenoviruses (AdVs) are non-enveloped icosahedral-shaped DNA viruses, causing infection in virtually all vertebrates. Their double-stranded linear DNA genomes range between 26 and 45kb in size, producing a broad repertoire of transcripts via highly complex alternative splicing patterns [@Davison2003; @Harrach2008]. The AdV genome is one of the most optimally economized; both the forward and reverse DNA strands harbor protein-coding genes, making it highly gene-dense. There are 16 genes termed "genus-common" that are homologous in all AdVs; these are thought to be inherited from a common ancestor. All other genes are termed "genus-specific". "Genus-specific" genes tend to be located at the termini of the genome while "genus-common" genes are usually central [@Davison2003]. This pattern is observed in *Adenoviridae*, *Poxviridae*, and *Herpesviridae* [@Davison2003; @Upton2003; @McGeoch1999]. The family *Adenoviridae* consists of five genera: *Mastadenovirus* (MAdV), *Aviadenovirus*, *Atadenovirus*, *Ichtadenovirus*, and *Siadenovirus* (SiAdV) [@Harrach2011; @Guimet2016]. Currently, there are three recognized members of the genus SiAdV: frog adenovirus 1, raptor adenovirus 1, and turkey adenovirus 3 also called turkey hemorrhagic enteritis virus (THEV) [@Harrach2011; @Kovács2011; @Davison2000; @Kovács2010; @Katoh2009]. Members of SiAdV have the smallest genome size (\~26 kb) and gene content (\~23 genes) of all known AdVs, and many "genus-specific" putative genes of unknown functions have been annotated (see **Figure 1**)[@Harrach2008; @Kovács2011; @Davison2003].

Virulent THEV strains (THEV-V) and avirulent strains (THEV-A) of THEV are serologically indistinguishable, infecting turkeys, chickens, and pheasants, with the THEV-V causing different clinical diseases in these birds [@Harrach2008; @Beach2006]. In turkeys, the THEV-V cause hemorrhagic enteritis (HE), a debilitating acute disease affecting predominantly 6-12-week-old turkeys characterized by immunosuppression (IS), weight loss, intestinal lesions leading to bloody diarrhea, splenomegaly, and up to 80% mortality [@Beach2006; @Beach2009a; @Gross1967]. HE is the most economically significant disease caused by any strain of THEV [@Beach2006]. While the current vaccine strain (a THEV-A isolated from a pheasant, Virginia Avirulent Strain [VAS]) has proven effective at preventing HE in young turkey poults, it still retains the immunosuppressive ability. Thus, vaccinated birds are rendered more susceptible to opportunistic infections and death than unvaccinated cohorts leading to substantial economic losses [@Beach2006; @Rautenschlein2000; @Larsen1985; @Dhama2017]. To eliminate this immunossupressive side-effect of the vaccine, a thorough investigation of the culprit viral factors (genes) mediating this phenomenon is essential. However, the transcriptome (splicing and gene expression patterns) of THEV has not been characterized, making the investigation of specific viral genes for possible roles in causing IS impractical. A well-characterized transcriptome of THEV is required to enable experimentation with specific viral genes that may mediate IS.

Myriads of studies have elucidated the AdV transcriptome in fine detail [@Donovan2020; @Zhao2014]. However, a large preponderance of studies focus on MAdVs -- specifically human AdVs. Thus, most of the current knowledge regarding AdV gene expression and replication is based on MAdV studies, which is generalized for all other AdVs [@Guimet2016; @Wolfrum2013]. MAdV genes are transcribed in a temporal manner; therefore, genes are categorized into five early transcription units (E1A, E1B, E2, E3, and E4), two intermediate (IM) units (pIX and IVa2), and one major late unit (MLTU or major late promoter [MLP] region), which generates five families of late mRNAs (L1-L5) based on the polyadenylation site. An additional gene (UXP or U exon) is located on the reverse strand. The early genes encode non-structural proteins such as enzymes or host cell modulating proteins, primarily involved in DNA replication, or providing the necessary intracellular niche for optimal replication while late genes encode structural proteins that act as capsid proteins, promote virion assembly, and direct genome packaging. The immediate early gene E1A is expressed first, followed by the the delayed early genes, E1B, E2, E3 and E4. Then the intermediate early genes, IVa2 and pIX are expressed followed by the late genes [@Zhao2014; @Donovan2020; @Guimet2016]. Noteworthily, the MLP shows basal transcriptional activity during early infection (before DNA replication), with a comparable efficiency to other early viral promoters, but reaches its maximal activity during late infection (after DNA replication). However, during early infection the repertoire of late transcripts from the MLP is restricted until late infection [@Guimet2016]. MAdV makes an extensive use of alternative RNA splicing to produce a very complex array of mRNAs. All but the pIX mRNA undergo at least one splicing event. For instance, the MLTU produces over 20 distinct splice variants all of which contain three non-coding exons at the 5'-end (collectively known as the tripartite leader, TPL) [@Donovan2020; @Zhao2014]. There is also an alternate 5' three non-coding exons present in varying amounts on a subset of MLTU mRNAs (known as the x-, y- and z-leaders). Lastly, there is the i-leader exon, which is infrequently included between the second and third TPL exons, and codes for the i-leader protein [@Falvey1983]. Thus, the MLTU produces a complex repertoire of mRNA with diverse 5' untranslated regions (UTRs) spliced onto different 3' coding exons which are grouped into five different 3'-end classes (L1-L5) based on polyadenylation site. Each transcription unit (TU) contains its own promoter driving the expression of all the array of mRNA transcripts produced via alternative splicing in the unit [@Zhao2014; @Donovan2020; @Guimet2016]. The promoters are activated at different phases of the infection by proteins from previously activated TUs. Paradoxically, the early-to-late phase transition during infection requires the L4 genes, 22K and 33K, which should only be available after the transition. However, a promoter in the L4 region (L4P) that directs the expression of these two proteins independent of the MLP was found, resolving the paradox [@Morris2010; @Guimet2016; @Donovan2020]. During translation of AdV mRNA, recent studies strongly suggest the potential usage of secondary start codons; adding to what was already a highly complex system for gene expression [@Donovan2020; @Westergren2021].

High throughput sequencing methods have facilitated the discovery of many novel transcribed regions and splicing isoforms. It is also a very powerful tool to study alternative splicing under different conditions at an unparalleled depth [@Djebali2012; @Zhao2014; Westergren2021]. In this paper, a paired-end deep sequencing experiment was performed to characterize for the first time the transcriptome of THEV (VAS vaccine strain) during different phases of the infection, yielding the first THEV splicing map. Our paired-end sequencing allowed for reading `r seq_len` bp long high quality (mean Phred Score of 36) sequences from each end of cDNA fragments, which were mapped to the genome of THEV.
\newpage

@@

**Overview of sequencing data and analysis pipeline outputs**  
A previous study by Aboezz *et al* showed that almost all THEV transcripts were detectable beginning at 4 hours [@Aboezz2019]. Therefore, infected MDTC-RP19 cells were harvested at 4-, 12-, 24-, and 72-hours post-infection (h.p.i) to ensure an amply wide time window to sample all transcripts. Our paired-end RNA sequencing (RNA-seq) experiment yielded an average of `r avg_tot_count` million total reads of `r seq_len`bp in length per time-point, which were simultaneously mapped to both the virus (THEV) and host (*Meleagris gallopavo*) genomes using the `Hisat2` [@Pertea2016] alignment program. A total of `r total_mapped` million reads from all time-points mapped to the virus genome; this provided good coverage/depth, leaving no regions unmapped. The mapped reads to the virus genome increased substantially from `r cov_4hr` reads at 4 h.p.i to `r cov_72hr` million reads at 72 h.p.i (**Table 1**, **Figure 2a**). From the mapped reads, we identified a total of `r total_unq_jncs_submit %>% nrow() %>% comma(.)` unique THEV splice junctions from all time-points, with splice junctions from the later time-points being supported by significantly more sequence reads than earlier time-points. For example, all the `r tot_jun_4` unique junctions at 4 h.p.i had less than 10 reads supporting each one, averaging a mere `r mean_juncreads_4` reads/junction. Conversely, the `r tot_jun_72` unique junctions at 72 h.p.i averaged `r mean_juncreads_72` reads/junction, some junctions having coverage as high as `r max_junread72` reads. The substantial increases in splice junction and mapping reads to the THEV genome over time denotes an active infection and correlates with our quantitative PCR (qPCR) assay quantifying the total number of viral genome copies over time (**Figure 2b**).

Using `StringTie` [@Pertea2016], an assembler of RNA-seq alignments into potential transcripts, the mapped reads for each time point were assembled into transcripts using the genomic location of the predicted THEV ORFs as a guide. In the consolidated transcriptome, a composite of all unredundant transcripts from all time points, we counted a total of `r nrow(spliced_gtf)-1` novel transcripts. Although some exons in some transcripts match the predicted ORFs exactly, most of our identified exons are longer, spanning multiple predicted ORFs (**Figure 3**).

We validated the splice junctions in all transcripts by PCR amplification of viral cDNA, cloning, and Sanger sequencing (**Supplementary PCR methods**). During validation, we identified 5 additional transcripts some of which were further validated by 3' Rapid Amplification of cDNA Ends (3'RACE) data. The complete list of unique splice junctions mapped to THEV's genome has been submitted to the National Center for Biotechnology Information Gene Expression Omnibus (<http://www.ncbi.nlm.nih.gov/geo>) under accession number GSE254416.

**Changes in THEV splicing profile over time**\
AdV gene expression occurs under exquisite temporal control with each promoter typically producing one or few pre-mRNAs that undergo alternative splicing to yield the manifold repertoire of complex transcripts. To evaluate the activity of each promoter over time, `StringTie` and `Ballgown` (a program for statistical analysis of assembled transcriptomes) [@Ballgown] were used to estimate the normalized expression levels of all transcripts for each time point in Fragments Per Kilobase of transcript per Million mapped reads (FPKM) units. Very few unique splice junctions, reads, and transcripts were counted at 4 h.p.i; hence, this time point was excluded in this analysis.

@@

Considering individual mRNAs, TRXPT_21 -- from the E2 region -- was the most significantly expressed at 12 h.p.i, constituting about `r abn_each_12`% of the total expression of all transcripts. Transcripts in the E3 and E4 regions also contributed significant proportions, and noticeably, some MLP region transcripts. The later time points were dominated by the MLP region transcripts -- TRXPT_10 and TRXPT_14 were the most abundantly expressed at 24 and 72 h.p.i, respectively, as expected (**Figure 4a**). When we performed analysis of the FPKM values of transcripts per region we found a similar pattern: the E2 region was the most abundantly expressed at 12 h.p.i, after which the MLP region assumes predominance (**Figure 4b**). Secondly, we estimated relative abundances of all splice junctions at each time point using the raw reads. For individual junctions, we counted as significantly expressed only junctions with coverage of at least 1% of the total splice junction reads at the given time point. At 12 h.p.i, `r nrow(sig_12_juncs)` junctions meet the 1% threshold, and were comprised of predominantly early region (E1, E2, E3, and E4) junctions, albeit the MLTU was the single most preponderant region overall, constituting `r round(most_sig_12, 1)`% of all the junction reads (**Table 2a** and **Supplementary Table S1a**). The topmost abundant junctions at 12 h.p.i remained the most significantly expressed at 24 h.p.i also. However, here, the MLP-derived junctions were unsurprisingly even more preponderant overall, accounting for `r round(most_sig_24, 1)`% of all the junction reads counted (**Table 2b** and **Supplementary Table S1b**). At 72 h.p.i, the trend of increased activity of the MLP continued as expected; at this time, the MLP region junctions were not only the most abundant overall -- accounting for `r round(most_sig_72, 1)`% of all junction reads, -- but also contained the most significantly expressed individual junctions (**Table 2c**, **Supplementary Table S1c** and **Figure 4c**). When we limited this analysis to only junctions in the final transcriptome, the relative abundances of the junctions for each region over time was generally similar to the pattern seen with all the junctions included (**Figure 4d**).

We also analyzed splice donor and acceptor site nucleotide usage over time to investigate any peculiarities that THEV may show, generally or over the course of the infection. We found that most splice donor-acceptor sequences were unsurprisingly the canonical GT-AG nucleotides. However, the splice acceptor-donor pairing became less specific over time, such that all combinations of nucleotide pairs were eventually detected (**Figure 5**)

**Early Region 1 (E1) transcripts**    
This region in MAdVs is the first transcribed after successful entry of the viral DNA into the host cell nucleus, albeit at low levels [@Zhao2014]. The host transcription machinery solely mediates the transcription of this region. After their translation, the E1 proteins in concert with a myriad of host transcription factors activate the other viral promoters [@Guimet2016]. In MAdVs, this region is subdivided into E1a and E2b units, but the transcripts found in our data categorized under this region do not appear to so divided.

>>>>

---
title: Characterizing the Transcriptome of Turkey Hemorrhagic Enteritis Virus in a Turkey B-cell Line
output:
  officedown::rdocx_document:
    reference_docx: style_word_output.docx
  pdf_document: default
geometry: margin=1in
font-size: 11pt
header-includes:
  - \usepackage{helvet}
  - \renewcommand*\familydefault{\sfdefault}
  - \usepackage{setspace}
  - \doublespacing
  - \usepackage[left]{lineno}
  - \linenumbers
  - \usepackage{booktabs}
  - \usepackage{longtable}
  - \usepackage{array}
  - \usepackage{multirow}
  - \usepackage{wrapfig}
  - \usepackage{float}
  - \usepackage{placeins}
  - \usepackage{colortbl}
  - \usepackage{pdflscape}
  - \usepackage{tabu}
  - \usepackage{threeparttable}
  - \usepackage{threeparttablex}
  - \usepackage[normalem]{ulem}
  - \usepackage{makecell}
  - \usepackage{xcolor}
  - \usepackage{caption}
  - \usepackage[labelformat=empty]{caption}
bibliography: transcriptome_refs.bib
csl: asm.csl
editor_options:
  chunck_output_type: console
---
\captionsetup[table]{labelformat=empty}

@@

Abraham Quaye^1\*^, Brett E. Pickett^\*^, Joel S. Griffitts^\*^, Bradford K. Berges^\*^, Brian D. Poole${^\dagger}$^\*^

@@

```{r call_libs, echo=FALSE,include=FALSE, message=FALSE, warning=FALSE}
library(knitr)
library(tidyverse)
library(tibble)
library(scales)
library(flextable)
library(english)
library(rticles)
opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
knitr::opts_chunk$set(tab.cap.pre = "", tab.cap.sep = "")
```

@@

Characterizing the splice map of turkey hemorrhagic enteritis virus (THEV) is an essential step that would allow studies of individual genes mediating its immunosuppressive functions. We used an RNA-sequencing experiment to characterize the transcriptome of THEV for the first time, providing key insight into the THEV gene expression and mRNA structures. Researchers previously annotated THEV's genome as encoding 23 open reading frames (ORFs). In this work we identified 29 spliced transcripts all of which consisted of novel exons although some exons matched some previously annotated ORFs. The three annotated splice junctions were also corroborated by our data. We performed PCR amplification of THEV cDNA, cloned the PCR products, and used Sanger sequencing to validate all identified splice junctions. During validation we identified five additional unique transcripts, a subset of which were further validated by 3' rapid amplification of cDNA ends (3' RACE) experiments. Thus, we report that the genome of THEV contains 34 unique transcripts with the coding capacity for all annotated ORFs. However, we found six of the previously annotated ORFs (ORF1, E3, 33K, ORF8, IVa2, and protease) to be truncated ORFs on the basis of the identification of an in-frame upstream start codon or the detection of additional coding exons. We also identified three of the annotated ORFs with longer or shorter isoforms, and seven novel unpredicted ORFs that could potentially be translated; although it is beyond the scope of this manuscript to investigate whether they are translated. Similar to other adenoviruses (AdVs), THEV also produces multiple distinctly spliced transcripts that code for the same proteins across its genome. Our data show that all THEV transcripts are spliced and organized into five transcription units under the control of their cognate promoters like other AdVs. However, our data suggest that the temporal regulation of THEV may be different from other AdVs.
\newpage

@@

Adenoviruses (AdVs) are non-enveloped icosahedral-shaped DNA viruses, causing infection in virtually all types of vertebrates studied to date. Their double-stranded linear DNA genomes range between 26 and 45kb in size, producing a broad repertoire of transcripts via highly complex alternative splicing patterns [@Davison2003; @Harrach2008]. The AdV genome is one of the most optimally economized; both the forward and reverse DNA strands harbor protein-coding genes, making it highly gene-dense. There are 16 genes termed "genus-common" that are homologous in all AdVs; these are thought to be inherited from a common ancestor. All other genes are termed "genus-specific". The genus-specific genes tend to be located at the termini of the genome while genus-common genes are usually towards the center of the genome [@Davison2003]. This pattern is also observed in *Poxviridae* and *Herpesviridae*, which also have linear DNA genomes [@Davison2003; @Upton2003; @McGeoch1999]. The family *Adenoviridae* consists of five genera: *Mastadenovirus* (MAdV), *Aviadenovirus*, *Atadenovirus*, *Ichtadenovirus*, and *Siadenovirus* (SiAdV) to which turkey adenovirus 3 also called turkey hemorrhagic enteritis virus (THEV) belongs [@Harrach2011; @Kovács2011; @Davison2000; @Kovács2010; @Katoh2009; @Guimet2016]. Members of SiAdV have the smallest genome size (\~26 kb) and gene content of all known AdVs (see **Figure 1**) [@Harrach2008; @Kovács2011; @Davison2003].

Virulent THEV strains (THEV-V) and avirulent strains (THEV-A) of THEV both infect turkeys, with THEV-V causing hemorrhagic enteritis (HE), a debilitating acute disease predominantly affecting turkey poults characterized by immunosuppression (IS), intestinal lesions leading to bloody diarrhea, and up to 80% mortality [@Beach2006; @Beach2009a; @Gross1967; @Harrach2008]. While the current vaccine strain (a THEV-A strain isolated from a pheasant; Virginia Avirulent Strain [VAS]) has proven effective at preventing HE in turkey poults, it still retains its immunosuppressive ability. Thus, vaccinated birds are rendered more susceptible to opportunistic infections and death than unvaccinated birds leading to substantial economic losses [@Beach2006; @Rautenschlein2000; @Larsen1985; @Dhama2017]. To eliminate the immunossupressive effect of the vaccine strain, a thorough investigation of the culprit viral genes mediating this phenomenon is essential. However, the transcriptome (splicing and gene expression patterns) of THEV has not been characterized, making an investigation of specific IS-related viral genes impractical.

A myriad of studies have elucidated the AdV transcriptome in fine detail [@Donovan2020; @Zhao2014]. However, a large preponderance of studies focus on MAdVs -- specifically human AdVs. Thus, most of the current knowledge regarding AdV gene expression and replication is based on MAdV studies, which is generalized for all other AdVs [@Guimet2016; @Wolfrum2013]. MAdV transcription is temporally regulated; therefore, genes are categorized into five early transcription units (E1A, E1B, E2, E3, and E4), two intermediate (IM) units (pIX and IVa2), and one major late transcription unit (MLTU or major late promoter [MLP] region), which generates five families of late mRNAs (L1-L5) based on the polyadenylation site. An additional gene (UXP or U exon) is located on the reverse strand. The early genes encode non-structural proteins such as enzymes or host-cell modulating proteins, primarily involved in DNA replication, or providing the necessary intracellular niche for optimal replication while late genes encode structural proteins that act as capsid proteins, promote virion assembly, or direct genome packaging. The immediate early genes E1A are expressed first, followed by the delayed early genes, E1B, E2, E3 and E4. Then the intermediate early genes, IVa2 and pIX are expressed followed by the late genes [@Zhao2014; @Donovan2020; @Guimet2016]. It is noteworthy that the MLP shows basal transcriptional activity during early infection (before DNA replication), with a comparable efficiency to other early viral promoters, but it reaches its maximal activity during late infection (after DNA replication). However, during early infection only a subset of the MLP-derived transcripts are expressed [@Guimet2016]. MAdV makes an extensive use of alternative RNA splicing to produce a very complex array of mRNAs. All but the pIX mRNA undergo at least one splicing event. For instance, the MLTU produces over 20 distinct splice variants all of which contain three non-coding exons at the 5'-end (collectively known as the tripartite leader; TPL) [@Donovan2020; @Zhao2014]. There is also an alternate three-exon 5' non-coding leader sequence present in varying amounts on a subset of MLTU mRNAs (known as the x-, y-, and z-leaders). Lastly, there is the i-leader exon, which is infrequently included between the second and third TPL exons, and codes for the i-leader protein [@Falvey1983]. Thus, the MLTU produces a complex repertoire of mRNA with diverse 5' untranslated regions (UTRs) spliced onto different 3' coding exons which are grouped into five different 3'-end classes (L1-L5) based on polyadenylation site. Each transcription unit (TU) contains its own promoter driving the expression of the array of mRNA transcripts produced via alternative splicing in the unit [@Zhao2014; @Donovan2020; @Guimet2016]. The promoters are activated at different phases of the infection by proteins from previously activated TUs. Paradoxically, the early-to-late phase transition during infection requires the L4 gene products, 22K and 33K, which should only be available after the transition. However, a promoter in the L4 region (L4P) that directs the expression of these two proteins independent of the MLP was found, resolving the paradox [@Morris2010; @Guimet2016; @Donovan2020]. During translation of AdV mRNA, recent studies using long-read direct RNA sequencing strongly suggest the potential usage of secondary start codons; adding to what was already a highly complex system for gene expression [@Donovan2020; @Westergren2021].

High throughput sequencing methods have facilitated the discovery of many novel transcribed regions and splicing isoforms. It is also a very powerful tool to study alternative splicing under different conditions at an unparalleled depth [@Djebali2012; @Zhao2014; @Westergren2021]. In this paper, we use a paired-end deep sequencing experiment to characterize, for the first time, the transcriptome and splicing of THEV (VAS vaccine strain) during different phases of the infection. Our paired-end sequencing allowed for reading `r seq_len` bp long high quality (mean Phred Score of 36) sequences from each end of cDNA fragments, which were mapped to the genome of THEV.
\newpage

@@

**Overview of sequencing data and analysis pipeline outputs**  
A previous study by Aboezz *et al* showed that almost all THEV transcripts were detectable beginning at 4 hours, with a full replication cycle of about 18-hours post infection (hpi) [@Aboezz2019]. Therefore, infected MDTC-RP19 cells were harvested at 4-, 12-, 24-, and 72-hpi  to ensure a wide time window to sample all transcripts. Our paired-end RNA sequencing (RNA-seq) experiment yielded an average of `r avg_tot_count` million total reads of `r seq_len` bp in length per time-point, which were simultaneously mapped to both the virus (THEV) and host (*Meleagris gallopavo*) genomes using the Hisat2 [@Pertea2016] reference-based aligner. A total of `r total_mapped` million reads from all time-points mapped to the virus genome; this provided good coverage/depth, leaving no regions unmapped. The mapped reads to the virus genome increased substantially from a mere `r cov_4hr` reads at 4 hpi to `r cov_72hr` million reads at 72 hpi (**Table 1**, **Figure 2A**). From the mapped reads, we identified a total of `r total_unq_jncs_submit %>% nrow() %>% comma(.)` unique THEV splice junctions from all time-points, with splice junctions from the later time-points being supported by significantly more sequence reads than earlier time-points. For example, all the `r tot_jun_4` unique junctions at 4 hpi had less than 10 reads supporting each one, averaging a mere `r mean_juncreads_4` reads per junction. Conversely, the `r tot_jun_72` unique junctions at 72 hpi averaged `r mean_juncreads_72` reads per junction, some junctions having coverage as high as `r max_junread72` reads. The substantial increases in splice junction and mapping reads to the THEV genome over time denotes an active infection and a successful viral replication, and correlates with our quantitative PCR (qPCR) assay quantifying the total number of viral genome copies over time (**Figure 2B**).

Using StringTie [@Pertea2016], we assembled the data into potential transcripts using the genomic location of the prior predicted THEV ORFs as a guide. In the consolidated transcriptome, a composite of all non-redundant transcripts from all time points, we counted a total of `r nrow(spliced_gtf)-1` novel transcripts. We found that a subset of exons in the viral transcripts match the predicted ORFs exactly, with the majority of the exons being longer and spanning multiple predicted ORFs (**Figure 3**).

We then validated the splice junctions in all transcripts by PCR amplification of viral cDNA, cloning, and Sanger sequencing (**Supplementary PCR methods**). During validation, we identified five additional transcripts, some of which were further validated by 3' Rapid Amplification of cDNA Ends (3' RACE) data. The complete list of unique splice junctions mapped to the THEV genome has been submitted to the National Center for Biotechnology Information Gene Expression Omnibus (<http://www.ncbi.nlm.nih.gov/geo>) under accession number GSE254416.

**Changes in THEV splicing profile over time**\
AdV gene expression occurs under exquisite temporal control with each promoter typically producing one or few pre-mRNAs that undergo alternative splicing to yield a repertoire of mature mRNAs. To evaluate the activity of each promoter over time, StringTie and Ballgown (a program for statistical analysis of assembled transcriptomes) [@Ballgown] were used to estimate the normalized expression levels of all transcripts for each time point in Fragments Per Kilobase of transcript per Million mapped reads (FPKM) units. Very few unique splice junctions, reads, and transcripts were counted at 4 hpi; hence, this time point was excluded in this analysis.

@@

Considering individual mRNAs, TRXPT_21 -- from the E2 region -- was the most significantly expressed at 12 hpi, constituting `r abn_each_12`% of the total expression of all transcripts. Transcripts in the E3 and E4 regions also contributed significant proportions, and noticeably, some MLP region transcripts. The later time points were dominated by the MLP region transcripts -- TRXPT_10 and TRXPT_14 were the most abundantly expressed at 24 and 72 hpi, respectively, as expected (**Figure 4A**). When we performed analysis of the FPKM values of transcripts per region we found a similar pattern: the E2 region was the most abundantly expressed at 12 hpi, after which the MLP region assumes predominance (**Figure 4B**). Secondly, we estimated relative abundances of all splice junctions at each time point using the raw reads. Only junctions with a read coverage of at least 1% of the total splice junction reads at the given time point were considered significant and included in **Tables 2a-2c**. At 12 hpi, `r nrow(sig_12_juncs)` junctions meet the 1% threshold, and were comprised of predominantly early region (E1, E2, E3, and E4) junctions, albeit the MLTU was the single most preponderant region overall, constituting `r round(most_sig_12, 1)`% of all the junction reads (**Table 2a** and **Supplementary Table S1a**). The most abundant junctions at 12 hpi remained the most significantly expressed at 24 hpi also. However, here, the MLP-derived junctions were unsurprisingly even more preponderant overall, accounting for `r round(most_sig_24, 1)`% of all the junction reads counted (**Table 2b** and **Supplementary Table S1b**). At 72 hpi, the trend of increased activity of the MLP continued as expected; at this time, the MLP region junctions were not only the most abundant overall -- accounting for `r round(most_sig_72, 1)`% of all junction reads, -- but also contained the most significantly expressed individual junctions (**Table 2c**, **Supplementary Table S1c** and **Figure 4C**). When we limited this analysis to only junctions in the final transcriptome, we observed the relative abundances of the junctions for each region over time to be similar to the pattern seen with all the junctions included (**Figure 4D**).

We also analyzed splice donor and acceptor site nucleotide usage over time to investigate any peculiarities that THEV may show, generally or over the course of the infection. We found that most splice donor-acceptor sequences were unsurprisingly the canonical GU-AG nucleotides. However, the splice acceptor-donor pairing became less specific over time, such that all combinations of nucleotide pairs were eventually detected (**Figure 5**).

**Early Region 1 (E1) transcripts**    
This region in MAdVs is the first transcribed after successful entry of the viral DNA into the host cell nucleus, albeit at low levels [@Zhao2014]. The host transcription machinery solely mediates the transcription of this region. After their translation, the E1 proteins in concert with a myriad of host transcription factors activate other viral promoters [@Guimet2016]. In MAdVs, this region is subdivided into E1a and E2b units, but the transcripts found in our data categorized under this region do not appear to be subdivided.

Only two ORFs (ORF1 [sialidase] and Hyd) are predicted in this region; however, we discovered `r as.character(english(nrow(filter(spliced_gtf, region == "E1"))))` novel transcripts in this region, which collectively contain `r english(nrow(reg_e1_brkdown))` unique splice junctions (**Figure 6**). Most of the ORFs of the novel transcripts are distinct from the predicted ORFs, but they all have the coding potential (CP) for the predicted Hyd protein as the 3'-most coding sequence (CDS) if secondary start codon usage is considered as reported for other AdVs [@Donovan2020; @Zhao2014]. The 5'-most CDS of TRXPT_1 is multi-exonic, encoding a novel 17.9 kilodalton (kDa), 160 residue [amino acids (aa)] protein (ORF9). From its 5'-most start codon (SC), TRXPT_2 encodes the largest protein in this region -- a 64.3 kDa, 580 aa protein (ORF10) with the same SC as ORF9 (position 211 bp). ORF10 spans almost the entire predicted ORF1 and Hyd, coming short in two regards: it is spliced from 1655 bp to 1964 bp (ORF1's C-terminus, including the stop codon), and its stop codon (STC; position 2312) is 13 bp short of the Hyd STC. However, it has an SC 102 bp upstream and in-frame with ORF1's predicted SC. Thus, ORF10 shares substantial protein sequence similarity with ORF1 but not with Hyd, as the SC of Hyd is not in-frame. Without its splice site removing the ORF1 STC, TRXPT_2 would encode a longer variant of ORF1, starting from an upstream SC. TRXPT_3 is almost identical to TRXPT_1, except for the lack of TRXPT_1's second exon. Our RNA-seq data show that all E1 transcripts share the same transcription termination site (TTS; at position 2325 bp). However, TRXPT_3 and TRXPT_4 seem to have transcription start sites (TSS) downstream of the TSS of TRXPT_1 and TRXPT_2 (E1 TSS; position: 54 bp). Given that studies in MAdVs show that E1 mRNAs share not only a common TTS but also the TSS, and only differ from each other regarding the internal splicing [@Zhao2014], it is likely that TRXPT_3 and TRXPT_4 are incomplete, and their actual TSS just like the TTS are identical for all E1 transcripts. Regardless of the TSS considered for TRXPT_3, the CP remains unaffected. Its 5'-most CDS, beginning at 1965 bp and sharing the same STC as ORF9, produces a 13.1 kDa, 115 residue protein (ORF4). ORF4 was predicted in an earlier study [@Pitcovski1998] but was excluded in later studies [@Beach2009a; @Davison2003]; however, our data suggest it is a genuinely expressed ORF. Unlike TRXPT_3, the CP of TRXPT_4 is affected by the TSS considered; if we consider its unmodified TSS, then its CP is identical to TRXPT_3. However, if we assume that TRXPT_4 uses the E1 TSS, then the 5'-most CDS is a distinct, novel, multi-exonic 15.9 kDa, 143 aa protein (ORF11) with the same SC as ORF9 and ORF10 but with a unique STC.

The splice junctions of all transcripts in this region (except the junction for TRXPT_4) were validated by cloning of viral cDNA and Sanger sequencing (**Supplementary PCR methods**). During the validation of TRXPT_2, we found ORF1 to be present on the agarose gel (an unspliced band size) and Sanger sequencing results showed it to be a transcribed mRNA (**Supplementary PCR methods**). This was corroborated by our 3' RACE experiment, which showed a transcript (TRXPT_2B) spanning the entire ORF1 and Hyd ORFs without any splicing, with a poly-A tail immediately after the E1 TTS. The 5'-most CDS of this transcript (TRXPT_2B) would encode ORF1. However, TRXPT_2B has an upstream and in-frame SC to the predicted SC of ORF1, suggesting that the predicted ORF1 CDS is truncated -- the actual ORF1 (eORF1) that is expressed shares the same SC as ORF10, but has a unique STC.

**Early Region 2 (E2) and Intermediate Region (IM) transcripts**     
The E2 TU expressed on the anti-sense strand is subdivided into E2A and E2B and encodes three classical AdV proteins --  pTP and Ad-pol (E2B proteins), and DBP (E2A protein) -- essential for genome replication [@Donovan2020; @Zhao2014]. Unlike MAdV where two promoters (E2-early and E2-late) are known [@Donovan2020], we discovered only a single TSS (E2 TSS; 18,751 bp) from which both E2A and E2B transcription is initiated. However, similar to MAdVs, E2A and E2B transcripts have distinct TTSs, and the E2B transcripts share the TTS of the IVa2 transcript of the IM region [@Donovan2020; @Zhao2014] (**Figure 7**).
